---
title: "Final Project Write Up"
author: "Jeremy Mitaux"
date: "2025-12-10"
output: html_document
---

## Introduction

Bees provide essential services for both wild and cultivated plants, yet global and regional declines in bee diversity are increasingly reported (Potts, Simon G., et al. 2016; Ollerton, 2017). In the Pacific Northwest, native bee conservation has become a major concern as variations in yearly precipitation, temperature patterns, and floral phenology continue to increase (Kerr, Jeremy T., et al. 2015). Understanding how climatic gradients influence bee distributions is critical for predicting future biodiversity shifts and identifying conservation priorities. Insights from such studies is invaluable to ecologists and governments as we adapt to a changing world. Oregon’s diverse habits from the wet Coast Range to the arid high desert, create natural gradients for testing how climate (and in particular precipitation) affects native bee communities.

Previous studies have shown that bee richness in a particular area can be attributed to the local environment whether that be altitude (Hoiss et al., 2012) or where floral resources are abundant and not limited by drought or excessive moisture (Classen et al. 2015). In North America multiple analyses have found increased species homogenization in agricultural landscapes (Ponisio, et al. 2025). The Oregon Bee Atlas directed by OSU Extension and launched in 2018, fills a key gap by providing species-level records of bees across Oregon. Each record includes coordinates, date, collector, and associated floral host, enabling multi-year spatial analyses of bee diversity and composition.

One area that is not present in the data set however is precipitation data. This 
information could be useful for research by correlating bee species ocurrance with 
region specific climate data. Bees as pollinators play key roles in the ecosystem, 
but they require the availability of resources. Rain is a major part of that, and 
different environements respond differently to yearly precipitation changes.  

This project aims to address concerns due to these changing conditions. By combining Oregon Bee Atlas data along with precipitation measurements provided by the PRISM Group as OSU, along with Oregon ecoregions, I hope to answer the following: 

  1. In what ways do Oregon ecoregions compare in bee species richness? 
  
  2. Do yearly precipitation patterns influence species richness and which
regions are most affected? How does variability changes between ecoregions when precipitation patterns change?

What are my hypotheses? 

  1. More urbanized regions such as the Willamette Valley will see lower species abundance compared to less developed areas. Under the null hypothesis there isn't any difference.
  2. It is known that bees are not the biggest fans of rain, but plants definitely are.
I am curious to see, given increases in yearly differences, how ecoregions compare. 
Are some more or less resistant to drought or wetter conditions? Under my null, 
yearly precipitation change has no effect.

## Data 
1. List 
- PRISM Climate Data (annual precipitation, 2018–2024)
URL: https://prism.oregonstate.edu/recent/ 
Source: PRISM Climate Group / Oregon State University
Description: 4-km and 800m gridded monthly precipitation (ppt) and average climate (fahrenheit) data in raster for the continental US. Data is available in multiple temporal periods (daily, monthly, and yearly). This file will modified to only include Oregon. In addition I will be changing the precipitation values into in inches. 

- Oregon Ecoregions 
URL: https://geohub.oregon.gov/datasets/oregon-geo::oregon-ecoregions/explore 
Source: State of Oregon 
Description: "Ecoregions denote areas of general similarity in ecosystems and in the type quality, and quantity of environmental resources. This map depicts revisions and subdivisions of ecoregions that was compiled at a relatively small scale (Omernik 1987)." This shape file will be used to make groups by region of bee occurances, furthermore we can compare precipitation by region as well. 

- Oregon State Bounardary
URL: https://geohub.oregon.gov/datasets/oregon-state-boundary/about 
Description: Shape file used to crop the precipitation raster data which includes the entire lower 48.

- Oregon Bee Atlas 
With the Oregon Bee Atlas Data, my plan is to get rid of the columns I don’t plan on using, leaving just the bee species identified and location. From there I will match the records with the PRISM data.
Citation: 
Best, L., J. Engler, C. Feuerborn, J. Larsen, B. Lindh, C. J. Marshall, A. Melathopoulos, S. Kincaid, S.V. J. Robinson. 2022. Oregon Bee Atlas: Wild bee findings from 2019. Catalog of the Oregon State Arthropod Collection. 6(1): 1—13. DOI: http://dx.doi.org/10.5399/osu/cat_osac.6.1.4906.

Data Skills Used:
- Manipulation of tables, time series filtering, functions 
- Raster and vector spacial data analysis 
- Regression as a method of measuring correlation between varaibles
- t-test and interpreting p-values 

2. Load in packages and data

```{r load-libraries-3, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
library(terra) 
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(sf) 
```

Import Oregon Bee Atlas Data, remove columns, and filter years 
```{r}
# Import OBA data, and filter the longitude and latitude columns 
oba_import <- read.csv("data/OBA_2018-2024.csv")
oba_clean <- oba_import %>%
  filter(!is.na(decimalLatitude),
         !is.na(decimalLongitude)) %>%
  filter(decimalLatitude  >= 41.99179  & decimalLatitude  <= 46.23576,
         decimalLongitude >= -124.5595 & decimalLongitude <= -116.4699)
oba_clean <- oba_clean[, c("year", "decimalLatitude", "decimalLongitude",
                           "genus", "specificEpithet")]

# Creating a spatial object 
oba_sf <- st_as_sf(
  oba_clean,
  coords = c("decimalLongitude", "decimalLatitude"),
  crs = 4269,      
  remove = FALSE
)
```

Load in PRISM data, shape file, and Oregon eco regions
```{r}
# Precipitation raster data 
rast_2018 <- rast("data/prism_2018/prism_ppt_us_30s_2018.tif")
rast_2019 <- rast("data/prism_2019/prism_ppt_us_30s_2019.tif")
rast_2020 <- rast("data/prism_2020/prism_ppt_us_30s_2020.tif")
rast_2021 <- rast("data/prism_2021/prism_ppt_us_30s_2021.tif")
rast_2022 <- rast("data/prism_2022/prism_ppt_us_30s_2022.tif")
rast_2023 <- rast("data/prism_2023/prism_ppt_us_30s_2023.tif")
rast_2024 <- rast("data/prism_2024/prism_ppt_us_30s_2024.tif")

or_bound <- vect("data/Oregon_Boundary/Oregon_State_Boundary.shp")
or_bound <- project(or_bound, rast_2024)  # match CRS of raster


rast_2018 <- crop(rast_2018, or_bound) # for 2018
rast_2018 <- mask(rast_2018, or_bound)
rast_2018$precip_inch <- rast_2018$prism_ppt_us_30s_2018 / 25.4 #convert mm to in 
precip_2018 <- as.data.frame(rast_2018, xy = TRUE) # make into a data frame

rast_2019 <- crop(rast_2019, or_bound) 
rast_2019 <- mask(rast_2019, or_bound)
rast_2019$precip_inch <- rast_2019$prism_ppt_us_30s_2019 / 25.4 
precip_2019 <- as.data.frame(rast_2019, xy = TRUE) 

rast_2020 <- crop(rast_2020, or_bound) 
rast_2020 <- mask(rast_2020, or_bound)
rast_2020$precip_inch <- rast_2020$prism_ppt_us_30s_2020 / 25.4 
precip_2020 <- as.data.frame(rast_2020, xy = TRUE) 

rast_2021 <- crop(rast_2021, or_bound) 
rast_2021 <- mask(rast_2021, or_bound)
rast_2021$precip_inch <- rast_2021$prism_ppt_us_30s_2021 / 25.4 
precip_2021 <- as.data.frame(rast_2021, xy = TRUE) 

rast_2022 <- crop(rast_2022, or_bound) 
rast_2022 <- mask(rast_2022, or_bound)
rast_2022$precip_inch <- rast_2022$prism_ppt_us_30s_2022 / 25.4 
precip_2022 <- as.data.frame(rast_2022, xy = TRUE) 

rast_2023 <- crop(rast_2023, or_bound) 
rast_2023 <- mask(rast_2023, or_bound)
rast_2023$precip_inch <- rast_2023$prism_ppt_us_30s_2023 / 25.4 
precip_2023 <- as.data.frame(rast_2023, xy = TRUE) 

rast_2024 <- crop(rast_2024, or_bound) 
rast_2024 <- mask(rast_2024, or_bound)
rast_2024$precip_inch <- rast_2024$prism_ppt_us_30s_2024 / 25.4 
precip_2024 <- as.data.frame(rast_2024, xy = TRUE) 

# Oregon eco region
or_ecoregion <- vect("data/oregon_ecoregion/ecoregion.shp") |> 
  st_as_sf()
or_ecoregion <- st_transform(or_ecoregion, 4269) %>%
  filter(LEV3_NAME != "Puget Lowland") # I have no idea why this was in the data but removing it because pretty sure that's in Washington
or_ecoregion <- or_ecoregion %>%
  mutate(
    LEV3_NAME = recode(
      LEV3_NAME,
      "Eastern Cascades Slopes and Foothills" = "Eastern Cascades" # easier to read
    )
  )
##
or_bound_sf <- st_as_sf(or_bound)
```

## Data Wrangling, Summarizing, and Exploring 

```{r}
# This will be the data used from the Oregon Bee Atlas
# I've removed all the columns that I don't plan on using, such as plants, collecting method, collector name, etc. 
# I have also made 'oba_sf' which is a shapefile that can be used to plot bee occurrences with the raster data 
head(oba_clean)
```

```{r}
# Joining each bee record to its respective Level 3 eco region  
oba_eco <- st_join(
  oba_sf,
  or_ecoregion[, c("LEV3_NAME")],  # just keep the ecoregion name
  join = st_intersects,
  left = TRUE
)

#This creates a metric for 'species richness' by counting unique strings
eco_rich <- oba_eco %>%
  st_drop_geometry() %>%
  group_by(LEV3_NAME) %>%
  filter(!is.na(LEV3_NAME)) %>% # remove any empty values
  summarize(
    richness = n_distinct(paste(genus, specificEpithet)),
    .groups = "drop"
  )
or_eco_rich <- or_ecoregion %>%
  left_join(eco_rich, by = "LEV3_NAME")
```

To calculate richness, the 'genus' and 'specificEpithet' columns are pasted together.
From there we calculate the number of unique strings with the n_distinct function. 
NOTE: This isn't a perfect measurement of species richness. This measurement only indicates how many different species were found, but doesn't indicate the abundance of them. Therefore there could be regions that are low in 'richness' yet still have high
populations of pollinators.

```{r}
eco_rich_yr <- oba_eco %>%
  st_drop_geometry() %>% 
  group_by(year, LEV3_NAME) %>% # filter by year
  filter(!is.na(LEV3_NAME)) %>% # remove any empty values
  summarize(
    richness = n_distinct(paste(genus, specificEpithet)),
    .groups = "drop"
)
eco_rich_yr
```

Lets make a visualization to better understand this

```{r}
eco_totals <- eco_rich_yr %>%
  group_by(LEV3_NAME) %>%
  summarise(total_richness = sum(richness, na.rm = TRUE))

# I wanted the chart to show the ecoregions in a west to east manner
eco_order <- c(
  "Coast Range",
  "Willamette Valley",
  "Cascades",
  "Eastern Cascades", 
  "Klamath Mountains",
  "Blue Mountains",
  "Northern Basin and Range",
  "Central Basin and Range",
  "Columbia Plateau",
  "Snake River Plain"
)
eco_rich_yr$LEV3_NAME <- factor(eco_rich_yr$LEV3_NAME, levels = eco_order)

ggplot(eco_rich_yr, aes(x = LEV3_NAME, y = richness)) +
  geom_col(aes(fill = factor(year)), position = "dodge") +
  labs(
    x = "Ecoregion",
    y = "# of unique strings\n(genus + species)",
    fill = "Year",
    title = "Species Richness by Ecoregion and Year)"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
\n I like this visualization for a couple reasons. For one its nice how it kind of reads from 
west to east on the x-axis. Furthermore each ecoregion has species abundance by year, so you can 
compare regions by region and year simultaneously.

A couple things I notice here is that although most regions show shrinking species abundance, 
regions such as the Willamette Valley could possibly be changing at a faster rate. 


```{r}
#Plot ecoregions and create fun custom colors for them all 
# I had an llm help with color choices and make edits to visualizations 
    # for example like how to center the tile on this chart and custom colors/themes
    # for that reason some of the code isn't that graceful, but hope the visualizations are appealing enough and most of all get the point across

eco_cols <- c(
  "Coast Range" = "#1b9e77",
  "Blue Mountains" = "#7570b3",
  "Willamette Valley" = "#e7298a",
  "Columbia Plateau" = "#66a61e",
  "Eastern Cascades Slopes and Foothills" = "#e6ab02",
  "Cascades" = "#a6761d",
  "Snake River Plain" = "#666666",
  "Northern Basin and Range" = "#1f78b4",
  "Klamath Mountains" = "#b2df8a",
  "Central Basin and Range" = "#fb9a99"
)
ggplot() +
  geom_sf(data = or_ecoregion, 
          aes(fill = LEV3_NAME), 
          color = NA) +
  theme_void() +
  labs(
    fill = "Level III Ecoregions",
    title = "Oregon Ecoregions") +
    theme(plot.title = element_text(hjust = .5))
```

```{r}
#plot species richness by eco region
ggplot() +
  geom_sf(
    data = or_eco_rich,
    aes(fill = richness),
    color = NA,
    size = 0.2
  ) +
  scale_fill_viridis_c(
    option = "plasma",
    na.value = "grey90",
    name = "Species richness"
  ) +
  theme_void() +
  labs(
    title = "Bee Species Richness by Ecoregion",
    subtitle = "# of distinct strings when combining genus and species",
    x = "Longitude",
    y = "Latitude",
  )
```
Now lets incorporate the precipitation data 

```{r}
# basic plot for 2024
ggplot() +
  geom_raster(data = precip_2024,
              aes(x = x, y = y, fill = precip_inch)
  ) +
labs(
    title = "Oregon Precipitation, 2024",
    x = "Longitude",
    y = "Latitude",
  )
```
Lets use this precipitation data to learn more from the OBA data. What I am going
to do now is create mean precipitation values for each ecoregion. 

```{r}
mean_precip_by_ecoregion <- function(r_precip, year) {
  # Get means for each polygon in ecoregion file
  vals <- terra::extract(
    r_precip,
    vect(or_ecoregion),
    fun   = mean,
    na.rm = TRUE)
  
   # Group these means by LEV3 
  mean_yr <- or_ecoregion %>%
    st_drop_geometry() %>%
    select(LEV3_NAME) %>%
    mutate(
      year        = year,
      mean_precip = vals$precip_inch
    ) %>%
  select(year, LEV3_NAME, mean_precip
    )
  mean_yr <- mean_yr %>%
  group_by(year, LEV3_NAME) %>%
  summarize(mean_precip = mean(mean_precip, na.rm = TRUE), .groups = "drop")
}
```

```{r}
#2018
mean_precip_2018 <- mean_precip_by_ecoregion(
  r_precip     = rast_2018,      
  year         = 2018)
#2019
mean_precip_2019 <- mean_precip_by_ecoregion(
  r_precip     = rast_2019,      
  year         = 2019)
#2020
mean_precip_2020 <- mean_precip_by_ecoregion(
  r_precip     = rast_2020,     
  year         = 2020)
#2021
mean_precip_2021 <- mean_precip_by_ecoregion(
  r_precip     = rast_2021,     
  year         = 2021)
#2022
mean_precip_2022 <- mean_precip_by_ecoregion(
  r_precip     = rast_2022,    
  year         = 2022)
#2023
mean_precip_2023 <- mean_precip_by_ecoregion(
  r_precip     = rast_2023,    
  year         = 2023)
#2024
mean_precip_2024 <- mean_precip_by_ecoregion(
  r_precip     = rast_2024,   
  year         = 2024)

mean_precip <- bind_rows(mean_precip_2018, mean_precip_2019, mean_precip_2020,
                        mean_precip_2021, mean_precip_2022, mean_precip_2023,
                        mean_precip_2024)
eco_rich_yr <- eco_rich_yr %>%
  left_join(mean_precip, by = c("year", "LEV3_NAME")) %>%
  filter(year  > 2017)
```

```{r}
ggplot(eco_rich_yr, aes(mean_precip, richness)) +
  geom_point(size = 2, color = "firebrick") +
  geom_smooth(method = "lm", se = TRUE, fill = "pink") +
  labs(
    title = "Richness vs. Mean Precipitation: All Regions",
    x = "Mean Precipitation in.",
    y = "Richness"
  ) +
  theme_minimal(base_size = 14)
```
\n As we can see in this plot, there doesn't seem to be a clear linear relationship 
between precipitation and species richness. However this has just been some quick 
exploration, so lets see what more we can learn.

## Hypothesis Testing 

##Question 1
In what ways do Oregon ecoregions compare in bee species richness? 

```{r}
ggplot(eco_rich, aes(x=LEV3_NAME, y= richness)) +
  geom_col(position = "dodge", fill = "darkgreen") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Richness Counts by Ecoregion")
```

When look at overall species counts by ecoregion, we can see that there isn't a large difference. 
Mountainous and remote areas such as the Blue Moutains and Eastern Cascades have the largest counts (399 and 386 respectively), and dry xeric regions have the lowest (Snake River Plain with 99, and Central Basin and Range with 131) 

Lets plot a linear regression line to visualize the overall trend of species richness by
region overtime. 

```{r}
ggplot(eco_rich_yr, aes(x = year, y = richness)) +
  geom_col(aes(fill = factor(LEV3_NAME)), position = "dodge") +
  labs(
    x = "Ecoregion",
    y = "# of unique strings\n(genus + species)",
    fill = "Year",
    title = "Species Richness by Ecoregion and Year)"
  ) +
  geom_smooth(method = "lm", se = TRUE, fill = "pink") +
  theme_bw()
```
Overall species abundance is falling, but are regions being affected differently?

```{r}
ggplot(eco_rich_yr[eco_rich_yr$LEV3_NAME == "Willamette Valley", ],
       aes(year, richness)) +
  geom_point(size = 2, color = "firebrick") +
  geom_smooth(method = "lm", se = TRUE, fill = "pink") +
  labs(
    title = "Richness by Year: Willamette Valley",
    x = "Year",
    y = "Richness"
  ) +
  theme_minimal(base_size = 14)
```

```{r}
ggplot(eco_rich_yr[eco_rich_yr$LEV3_NAME == "Eastern Cascades", ],
       aes(year, richness)) +
  geom_point(size = 2, color = "firebrick") +
  geom_smooth(method = "lm", se = TRUE, fill = "pink") +
  labs(
    title = "Richness by Year: Eastern Cascades",
    x = "Year",
    y = "Richness"
  ) +
  theme_minimal(base_size = 14)
```

```{r}
ggplot(eco_rich_yr[eco_rich_yr$LEV3_NAME == "Columbia Plateau", ],
       aes(year, richness)) +
  geom_point(size = 2, color = "firebrick") +
  geom_smooth(method = "lm", se = TRUE, fill = "pink") +
  labs(
    title = "Richness by Year: Columbia Plateau",
    x = "Year",
    y = "Richness"
  ) +
  theme_minimal(base_size = 14)
```
From this exploration, there isn't a meaningful difference in Oregon ecoregions when looking at species richness. Going into this project/question I was expecting the Willamette Valley to have lower species richness, however with 260 species, this put its right at the average for Oregon.

## Question 2
Does yearly precipitation influence species richness and which
regions are most affected? How does variability changes between ecoregions when precipitation patterns change? Under my null there is not difference, lets test this. 

To begin, we'll run a quick linear model 

```{r}
mod_precip <- lm(richness ~ mean_precip, data = eco_rich_yr)
summary(mod_precip)
```
Here I fitted a linear model with richness as the response and mean precipitation as the predictor. The estimated slope was β₁ = 0.514. Because the p-value was 0.6087, I failed to reject my null hypothesis. This suggests that, within this dataset, precipitation doesn't have much of a large effect on species richness. That is if we are looking at Oregon as a whole. 

Lets focus on a specific region, in this case the Willamette Valley which is the most populated in Oregon. 

```{r}
ggplot(eco_rich_yr[eco_rich_yr$LEV3_NAME == "Willamette Valley", ]
       , aes(mean_precip, richness)) +
  geom_point(size = 2, color = "firebrick") +
  geom_smooth(method = "lm", se = TRUE, fill = "pink") +
  labs(
    title = "Richness vs. Mean Precipitation: Willamette Valley",
    x = "Mean Precipitation in.",
    y = "Richness"
  ) +
  theme_minimal(base_size = 14)
```

```{r}
mod_precip <- lm(richness ~ mean_precip, data = eco_rich_yr[eco_rich_yr$LEV3_NAME == "Willamette Valley", ])
summary(mod_precip)
```
Here the estimated effect are stronger (β = −4.94, SE = 2.59), suggesting that wetter years were associated with slightly lower species richness. However this effect was not statistically significant (t = −1.91, p = 0.1145), and I therefore fail to reject the null hypothesis.

Despite the non-significant result, the model explained a substantial portion of the variation (R² = 0.42), indicating a potentially meaningful relationship that may be obscured by small sample size. Overall, the evidence isn't strong enough to conclude that precipitation alone predicts richness in the Willamette Valley.

```{r}
ggplot(eco_rich_yr[eco_rich_yr$LEV3_NAME == "Willamette Valley", ]
       , aes(year, richness)) +
  geom_point(size = 2, color = "firebrick") +
  geom_smooth(method = "lm", se = TRUE, fill = "pink") +
  labs(
    title = "Richness by Year: Willamette Valley",
    x = "Year",
    y = "Richness"
  ) +
  theme_minimal(base_size = 14)
```
\n An even stronger association is richness by year.

```{r}
mod_interaction <- lm(richness ~ mean_precip * LEV3_NAME, data = eco_rich_yr)
anova(mod_interaction)
```

I tested whether the effect of precipitation on species richness differed among ecoregions using an interaction model. ANOVA results showed that mean precipitation did not significantly predict richness overall (F = 0.61, p = 0.44). Ecoregions naturally differed strongly in their average richness (F = 10.96, p = very small). 

The interaction between precipitation and ecoregion was not significant (F = 0.82, p = 0.60), meaning that precipitation again didn't have an effect across ecoregions. This suggests that while ecoregions can vary greatly in richness, these differences are not explained by precipitation alone, and precipitation does not appear to be a major driver of richness patterns in any particular region.

Overall, based the regression model across Oregon and the above interaction test, precipitation did not significantly predict richness. Results favor the null hypothesis. 

## Visualizations 

Lets make some nice visualizations of Oregon, with each ecoregion marketed by its mean 
precipitation. These aren't ground breaking, but they look nice and are good in tandem with
the previous visualizations made. 

```{r}
ggplot() +
  geom_sf(
    data = or_eco_rich,
    aes(fill = richness),
    color = NA,
    size = 0.2
  ) +
  geom_point(data = oba_sf,
             aes(x = decimalLongitude, y = decimalLatitude),
             size = 0.1,
             alpha = 0.1,
             color = "green"
  ) +
  scale_fill_viridis_c(
    option = "plasma",
    na.value = "grey90",
    name = "Species richness"
  ) +
  theme_void() +
  labs(
    title = "Bee Species Richness by Ecoregion",
    subtitle = "# of distinct strings when combining genus and species",
    x = "Longitude",
    y = "Latitude",
  )
```

```{r}
ggplot() +
  geom_raster(data = precip_2024,
              aes(x = x, y = y, fill = precip_inch)
  ) +
  geom_point(data = oba_sf[oba_sf$year == "2024", ],
             aes(x = decimalLongitude, y = decimalLatitude),
             size = 0.1,
             alpha = 0.1,
             color = "green"
  ) +
  scale_fill_viridis_c(
    option = "plasma",
    na.value = "grey90",
    name = "Precipitation in."
  ) +
  theme_void() +
  labs(
    title = "2024 Mean Precipitation with 2024 Bee Occurances",
    subtitle = "# of distinct strings when combining genus and species",
    x = "Longitude",
    y = "Latitude",
  )
```

# Conclusion and Further Thoughts

After analyzing the OBA data set with the inclusion of precipitation and ecoregion data, we can 
confidently conclude that Oregon's ecoregions follow similar patters in species richness, and that precipitation doesn't play a pivotal role in richness either. We failed to reject both our null hypothesis. 

A major point of note for this project is the species richness metric, which was calculated by counting the number of unique strings in a given ecoregion. These strings were created by pasting the genus and specificEpithet columns. Last friday when presenting my poster, I was lucky to have 
Lincoln Best in the crowd, who performs much of the taxonomic work for the atlas. He made a enlightening comment that completely re-frames the reality of my project. 

Simply put, much of the data from recent years is still yet to be finished, and there are many bees that are yet to be identified. Therefore  newer years are under reported compared to older ones. With this new info, we can argue that the negative relationships seen with species richness is likely due to this lapse in metadata. 

Lets do a diagnostic plot to see how well out linear model performed for question 2
```{r}
plot(mod_precip)   # Residuals, Q-Q plot
```

The diagnostic plots above indicate that several assumptions of linear regression are violated. The Residuals vs Fitted plot shows nonlinearity, while the Q–Q plot demonstrates deviations from normality, particularly in the tails. The Scale–Location plot reveals heteroscedasticity, and the Residuals vs Leverage plot identifies multiple points that strongly influence the regression line.

Taken together, these diagnostics indicate that a simple linear model hasn't been the appropriate representation of the relationship between precipitation and species richness. With the current setup of 7 years, it doesn't take much for the model to get thrown off. Therefore the nonsignificant relationship (p = 0.61) should be interpreted with weariness. Nevertheless, the absence of any clear trend in the data suggests that precipitation is very unlikely to be a meaningful predictor of richness.


